---
title: "S&P 500 — Market Risk Analysis"
author: "Karol Kudłacik"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.align = "center",
  fig.width = 9,
  fig.height = 5.5
)
```

*Historical VaR · Expected Shortfall (97.5%) · simplified IMA-style charge ·
Basel traffic-light backtesting with formal coverage tests.*

## Objective

Quantify the downside market risk of the S&P 500 and, crucially, **check whether
a standard 99% Value-at-Risk model actually holds up out-of-sample** — the
validation question a risk function has to answer before relying on a model for
capital. The pipeline goes from raw prices to regulatory-style metrics:
Historical VaR, Expected Shortfall, a (deliberately simplified) FRTB IMA-style
charge, and a Basel backtest of the model on the *realised* index returns.

**Data.** Daily S&P 500 OHLC prices, file `spx_d.csv` (source:
[Stooq](https://stooq.pl/q/d/l/?s=^spx&i=d); schema `Date,Open,High,Low,Close,Volume`,
only `Date`/`Open`/`Close` used), sample from 1995-01-01 onwards.

**Reproducibility.** Everything on the real index is deterministic. The only
stochastic part is *Appendix A* — a synthetic sanity-check of the traffic-light
classifier (not a result for the S&P 500) — seeded with `set.seed(42)`. The
analysis uses **base R only**; `knitr` is used here purely to render the tables.

## 1 Data loading & preparation

Prices are sorted, filtered from 1995, and reduced to the columns we need.

```{r load-data}
sp500_data       <- read.csv("spx_d.csv")
sp500_data$Date  <- as.Date(sp500_data$Date)
sp500_data       <- sp500_data[order(sp500_data$Date), ]

cutoff_date      <- as.Date("1995-01-01")
sp500_filtered   <- sp500_data[sp500_data$Date >= cutoff_date, c("Date", "Open", "Close")]
sp500_filtered   <- sp500_filtered[complete.cases(sp500_filtered), ]

close_vec        <- sp500_filtered$Close
dates_vec        <- sp500_filtered$Date

cat("Observations after 1995-01-01:", nrow(sp500_filtered), "\n")
cat("Date range:", format(min(dates_vec)), "to", format(max(dates_vec)), "\n")
```

## 2 Return computation

The simple (arithmetic) return over a holding period $h$ is
$r_t^{(h)} = (P_t - P_{t-h})/P_{t-h}$, with $h=1$ for daily and $h=10$ for the
ten-trading-day Basel horizon.

> **Note on the 10-day series.** These are *overlapping* 10-day returns (each
> shares 9 days with its neighbour). They are convenient but serially correlated
> by construction — which matters for exception counting and is flagged again
> below.

```{r returns}
returns_1d  <- diff(close_vec, lag = 1)  / head(close_vec, -1)
returns_10d <- diff(close_vec, lag = 10) / head(close_vec, -10)

dates_1d    <- dates_vec[-1]
dates_10d   <- dates_vec[-(1:10)]

cat("1-day  return observations:", length(returns_1d),  "\n")
cat("10-day return observations:", length(returns_10d), "\n")
```

## 3 Historical Value-at-Risk (99%)

A rolling 252-day Historical VaR at 99% confidence, using the $(n+1)\,p$
plotting-position quantile (R's `quantile` type 6). With $n=252$ and $p=0.01$
the tail position is $h=(252+1)\times 0.01 = 2.53$, so

$$\text{VaR}_t = -\Big[\,r_{(2)} + 0.53\,\big(r_{(3)} - r_{(2)}\big)\Big],$$

where $r_{(k)}$ is the $k$-th smallest return in the window. VaR is reported as a
**positive** loss number.

> **Alignment.** $\text{VaR}_t$ is estimated from returns $[t-252,\ t-1]$ and
> never uses day $t$ itself, so the estimate is genuinely out-of-sample. The
> *same* alignment is reused everywhere below (including the backtest), so no
> separate "shift" step is needed.

```{r var-helper}
hist_VaR <- function(returns_vec, window = 252, conf = 0.99) {
  n   <- length(returns_vec)
  p   <- 1 - conf                              # 0.01 for 99% VaR
  VaR <- rep(NA_real_, n)

  for (t in (window + 1):n) {
    window_returns <- returns_vec[(t - window):(t - 1)]   # 252 past returns
    sorted         <- sort(window_returns)                # ascending (worst first)

    h    <- (window + 1) * p                   # = 2.53
    lo   <- floor(h)                           # = 2
    hi   <- lo + 1                             # = 3
    frac <- h - lo                             # = 0.53

    VaR[t] <- -(sorted[lo] + frac * (sorted[hi] - sorted[lo]))
  }
  VaR
}
```

```{r var-compute}
VaR_1d  <- hist_VaR(returns_1d,  window = 252, conf = 0.99)
VaR_10d <- hist_VaR(returns_10d, window = 252, conf = 0.99)

cat("=== 1-Day Historical VaR (99%) ===\n");  print(summary(na.omit(VaR_1d)))
cat("\n=== 10-Day Historical VaR (99%) ===\n"); print(summary(na.omit(VaR_10d)))
```

```{r var-plots, fig.height=7}
par(mfrow = c(2, 2), mar = c(3, 4, 3, 1))

plot(dates_1d, returns_1d, type = "l", col = "steelblue", lwd = 0.7,
     main = "S&P 500 - 1-Day Simple Returns", xlab = "", ylab = "Return", xaxt = "n")
axis.Date(1, at = seq(min(dates_1d), max(dates_1d), by = "5 years"), format = "%Y")
abline(h = 0, col = "black", lwd = 0.8)

plot(dates_10d, returns_10d, type = "l", col = "darkorange", lwd = 0.7,
     main = "S&P 500 - 10-Day Simple Returns", xlab = "", ylab = "Return", xaxt = "n")
axis.Date(1, at = seq(min(dates_10d), max(dates_10d), by = "5 years"), format = "%Y")
abline(h = 0, col = "black", lwd = 0.8)

plot(dates_1d, returns_1d, type = "l", col = "steelblue", lwd = 0.7,
     main = "1-Day Returns vs. Rolling 99% VaR", xlab = "", ylab = "Return / VaR", xaxt = "n")
axis.Date(1, at = seq(min(dates_1d), max(dates_1d), by = "5 years"), format = "%Y")
lines(dates_1d, -VaR_1d, col = "firebrick", lwd = 1.5)
abline(h = 0, col = "black", lwd = 0.8)
legend("bottomleft", legend = c("1-day return", "-VaR 99%"),
       col = c("steelblue", "firebrick"), lwd = c(0.7, 1.5), bty = "n", cex = 0.8)

plot(dates_10d, returns_10d, type = "l", col = "darkorange", lwd = 0.7,
     main = "10-Day Returns vs. Rolling 99% VaR", xlab = "", ylab = "Return / VaR", xaxt = "n")
axis.Date(1, at = seq(min(dates_10d), max(dates_10d), by = "5 years"), format = "%Y")
lines(dates_10d, -VaR_10d, col = "firebrick", lwd = 1.5)
abline(h = 0, col = "black", lwd = 0.8)
legend("bottomleft", legend = c("10-day return", "-VaR 99%"),
       col = c("darkorange", "firebrick"), lwd = c(0.7, 1.5), bty = "n", cex = 0.8)

par(mfrow = c(1, 1))
```

### VaR exceedance counts (descriptive)

A 99% VaR is expected to be breached on ~1% of days. Formal coverage tests are
run on the 1-day series in the [backtesting section](#7-basel-backtesting-real-sp-500).

```{r var-exceedance}
breaches_1d  <- sum(-returns_1d  > VaR_1d,  na.rm = TRUE)
breaches_10d <- sum(-returns_10d > VaR_10d, na.rm = TRUE)
total_1d     <- sum(!is.na(VaR_1d))
total_10d    <- sum(!is.na(VaR_10d))

cat(sprintf("1-Day  VaR exceptions: %d / %d  (%.2f%%)  | expected 1.00%%\n",
            breaches_1d, total_1d, 100 * breaches_1d / total_1d))
cat(sprintf("10-Day VaR exceptions: %d / %d  (%.2f%%)  | overlapping returns\n",
            breaches_10d, total_10d, 100 * breaches_10d / total_10d))
```

The 10-day exceptions are serially dependent (overlap), so the count **cannot**
be compared to 1% the way the 1-day count can — treat it as descriptive only.

## 4 Expected Shortfall (97.5%)

Expected Shortfall answers: *given that we are in the worst 2.5% of days, what is
the average loss?* The 97.5% level is the FRTB convention. With $252\times0.025 =
6.3$ observations in the tail (6 fully + 0.3 weight on the 7th):

$$\text{ES}_t = -\frac{1}{6.3}\left[\sum_{k=1}^{6} r_{(k)} + 0.3\,r_{(7)}\right].$$

> **Convention note.** VaR above uses the $(n+1)\,p$ position; this ES uses the
> $n\,p$ tail mass ($252\times0.025=6.3$). They are deliberately different
> estimators — a quantile vs. a tail average.

```{r es-helper}
hist_ES <- function(returns_vec, window = 252, conf = 0.975) {
  n   <- length(returns_vec)
  p   <- 1 - conf                          # 0.025
  ES  <- rep(NA_real_, n)

  tail_n    <- window * p                  # 6.3
  tail_full <- floor(tail_n)               # 6
  tail_frac <- tail_n - tail_full          # 0.3

  for (t in (window + 1):n) {
    window_returns <- returns_vec[(t - window):(t - 1)]
    sorted         <- sort(window_returns)
    tail_sum <- sum(sorted[1:tail_full]) + tail_frac * sorted[tail_full + 1]
    ES[t]    <- -tail_sum / tail_n
  }
  ES
}
```

```{r es-compute}
ES_1d  <- hist_ES(returns_1d,  window = 252, conf = 0.975)
ES_10d <- hist_ES(returns_10d, window = 252, conf = 0.975)

cat("=== 1-Day Expected Shortfall (97.5%) ===\n");  print(summary(na.omit(ES_1d)))
cat("\n=== 10-Day Expected Shortfall (97.5%) ===\n"); print(summary(na.omit(ES_10d)))
```

```{r es-plots, fig.height=7}
par(mfrow = c(2, 1), mar = c(3, 4, 3, 1))

plot(dates_1d, returns_1d, type = "l", col = "steelblue", lwd = 0.6,
     main = "S&P 500 - 1-Day Returns | Rolling 99% VaR & 97.5% ES",
     xlab = "", ylab = "Return / Risk", xaxt = "n",
     ylim = range(c(returns_1d, -VaR_1d, -ES_1d), na.rm = TRUE))
axis.Date(1, at = seq(min(dates_1d), max(dates_1d), by = "5 years"), format = "%Y")
lines(dates_1d, -VaR_1d, col = "firebrick",  lwd = 1.5)
lines(dates_1d, -ES_1d,  col = "darkorchid", lwd = 1.5)
abline(h = 0, col = "black", lwd = 0.5)
legend("bottomleft", legend = c("1-day return", "-VaR 99%", "-ES 97.5%"),
       col = c("steelblue", "firebrick", "darkorchid"), lwd = c(0.6, 1.5, 1.5),
       bty = "n", cex = 0.8)

plot(dates_10d, returns_10d, type = "l", col = "darkorange", lwd = 0.6,
     main = "S&P 500 - 10-Day Returns | Rolling 99% VaR & 97.5% ES",
     xlab = "", ylab = "Return / Risk", xaxt = "n",
     ylim = range(c(returns_10d, -VaR_10d, -ES_10d), na.rm = TRUE))
axis.Date(1, at = seq(min(dates_10d), max(dates_10d), by = "5 years"), format = "%Y")
lines(dates_10d, -VaR_10d, col = "firebrick",  lwd = 1.5)
lines(dates_10d, -ES_10d,  col = "darkorchid", lwd = 1.5)
abline(h = 0, col = "black", lwd = 0.5)
legend("bottomleft", legend = c("10-day return", "-VaR 99%", "-ES 97.5%"),
       col = c("darkorange", "firebrick", "darkorchid"), lwd = c(0.6, 1.5, 1.5),
       bty = "n", cex = 0.8)

par(mfrow = c(1, 1))
```

### ES exceedance count — diagnostic only

Counting how often the loss exceeds the ES level is a monitoring **diagnostic,
not a validation of ES**. ES is not elicitable, so there is no clean "expected
frequency" to test against; a proper ES backtest would use e.g. the
Acerbi–Székely (2014) tests.

```{r es-exceedance}
exc_ES_1d  <- sum(-returns_1d  > ES_1d,  na.rm = TRUE)
exc_ES_10d <- sum(-returns_10d > ES_10d, na.rm = TRUE)

cat(sprintf("1-Day  ES exceedances: %d / %d  (%.2f%%)\n",
            exc_ES_1d, sum(!is.na(ES_1d)), 100 * exc_ES_1d / sum(!is.na(ES_1d))))
cat(sprintf("10-Day ES exceedances: %d / %d  (%.2f%%)\n",
            exc_ES_10d, sum(!is.na(ES_10d)), 100 * exc_ES_10d / sum(!is.na(ES_10d))))
```

## 5 VaR vs. ES comparison

```{r comparison-table}
comp <- data.frame(
  Metric   = c("VaR 99% - mean", "VaR 99% - max",
               "ES 97.5% - mean", "ES 97.5% - max", "ES / VaR ratio (avg)"),
  `1-Day`  = c(mean(VaR_1d, na.rm = TRUE), max(VaR_1d, na.rm = TRUE),
               mean(ES_1d, na.rm = TRUE),  max(ES_1d, na.rm = TRUE),
               mean(ES_1d / VaR_1d, na.rm = TRUE)),
  `10-Day` = c(mean(VaR_10d, na.rm = TRUE), max(VaR_10d, na.rm = TRUE),
               mean(ES_10d, na.rm = TRUE),  max(ES_10d, na.rm = TRUE),
               mean(ES_10d / VaR_10d, na.rm = TRUE)),
  check.names = FALSE
)
knitr::kable(comp, digits = 4, caption = "Rolling VaR and ES, full sample.")
```

**`√10` cross-check.** The regulatory shortcut for a 10-day horizon scales the
1-day VaR by $\sqrt{10}$ under an i.i.d. assumption. Comparing it to the direct
(overlapping) 10-day VaR is a sanity check, not a claim that one is correct — a
gap is expected because real returns are neither i.i.d. nor free of serial
correlation.

```{r sqrt-scaling}
cat(sprintf("Direct overlapping 10-day VaR (mean): %.4f\n", mean(VaR_10d, na.rm = TRUE)))
cat(sprintf("sqrt(10) * 1-day VaR (mean)         : %.4f\n", sqrt(10) * mean(VaR_1d, na.rm = TRUE)))
```

## 6 Simplified IMA-style charge

This captures the **spirit** of the FRTB Internal Models Approach — capital
driven by the larger of a stressed-period ES and a current ES at 97.5%:

$$\text{Charge} = \max\!\big(\text{ES}_{\text{stress}},\ \text{ES}_{\text{recent}}\big).$$

> **It is a deliberate simplification, not the regulatory IMA number.** It omits
> the capital multiplier ($m_c \ge 1.5$), liquidity-horizon scaling, the
> reduced-set calibration, the non-modellable-risk-factor (NMRF/SES) add-on, the
> default-risk charge (DRC), and cross-risk-class aggregation.

The stress window is found by scanning **every** 252-day window for the highest
ES; recent ES is the last 252 days.

```{r ima-setup}
window    <- 252
conf      <- 0.975
p         <- 1 - conf
tail_n    <- window * p          # 6.3 (globals used by es_single_window below)
tail_full <- floor(tail_n)       # 6
tail_frac <- tail_n - tail_full  # 0.3

es_single_window <- function(ret_window) {
  sorted   <- sort(ret_window)
  tail_sum <- sum(sorted[1:tail_full]) + tail_frac * sorted[tail_full + 1]
  -tail_sum / tail_n
}

find_stress_period <- function(returns_vec, dates_vec, window = 252) {
  n         <- length(returns_vec)
  n_windows <- n - window + 1
  es_all    <- numeric(n_windows)
  for (i in seq_len(n_windows)) {
    es_all[i] <- es_single_window(returns_vec[i:(i + window - 1)])
  }
  best_i <- which.max(es_all)
  list(stress_ES = es_all[best_i],
       stress_start = dates_vec[best_i],
       stress_end   = dates_vec[best_i + window - 1],
       stress_index = best_i,
       all_es       = es_all)
}

stress_1d  <- find_stress_period(returns_1d,  dates_1d)
stress_10d <- find_stress_period(returns_10d, dates_10d)

recent_ES_1d  <- es_single_window(tail(returns_1d,  window))
recent_ES_10d <- es_single_window(tail(returns_10d, window))

IMA_1d  <- max(stress_1d$stress_ES,  recent_ES_1d)
IMA_10d <- max(stress_10d$stress_ES, recent_ES_10d)
```

```{r ima-stress-table}
stress_tbl <- data.frame(
  Horizon            = c("1-day", "10-day"),
  `Stress window`    = c(paste(format(stress_1d$stress_start),  "to", format(stress_1d$stress_end)),
                         paste(format(stress_10d$stress_start), "to", format(stress_10d$stress_end))),
  `Stress ES (97.5%)`= c(stress_1d$stress_ES, stress_10d$stress_ES),
  check.names = FALSE
)
knitr::kable(stress_tbl, digits = 4, caption = "Auto-identified stress period (highest-ES 252-day window).")
```

```{r ima-charge-table}
ima_tbl <- data.frame(
  Metric   = c("Stress ES (97.5%)", "Recent ES (97.5%)",
               "Simplified IMA-style charge", "Binding constraint"),
  `1-Day`  = c(sprintf("%.4f", stress_1d$stress_ES), sprintf("%.4f", recent_ES_1d),
               sprintf("%.4f", IMA_1d),
               ifelse(stress_1d$stress_ES  >= recent_ES_1d,  "STRESS", "RECENT")),
  `10-Day` = c(sprintf("%.4f", stress_10d$stress_ES), sprintf("%.4f", recent_ES_10d),
               sprintf("%.4f", IMA_10d),
               ifelse(stress_10d$stress_ES >= recent_ES_10d, "STRESS", "RECENT")),
  check.names = FALSE
)
knitr::kable(ima_tbl, caption = "Simplified IMA-style charge = max(stress ES, recent ES).")
```

```{r stress-plots, fig.height=7}
par(mfrow = c(2, 2), mar = c(3, 4, 3, 1))

dates_scan_1d  <- dates_1d [1:(length(returns_1d)  - window + 1)]
dates_scan_10d <- dates_10d[1:(length(returns_10d) - window + 1)]

plot(dates_scan_1d, stress_1d$all_es, type = "l", col = "steelblue", lwd = 1,
     main = "ES of Every 252-Day Window - 1-Day", xlab = "", ylab = "ES (97.5%)", xaxt = "n")
axis.Date(1, at = seq(min(dates_scan_1d), max(dates_scan_1d), by = "5 years"), format = "%Y")
abline(v = stress_1d$stress_start, col = "firebrick",  lwd = 2,   lty = 2)
abline(h = stress_1d$stress_ES,    col = "firebrick",  lwd = 1.5, lty = 3)
abline(h = recent_ES_1d,           col = "darkorchid", lwd = 1.5, lty = 3)
legend("topright", legend = c("Rolling ES", "Stress start", "Stress ES", "Recent ES"),
       col = c("steelblue","firebrick","firebrick","darkorchid"),
       lwd = c(1,2,1.5,1.5), lty = c(1,2,3,3), bty = "n", cex = 0.75)

plot(dates_scan_10d, stress_10d$all_es, type = "l", col = "darkorange", lwd = 1,
     main = "ES of Every 252-Day Window - 10-Day", xlab = "", ylab = "ES (97.5%)", xaxt = "n")
axis.Date(1, at = seq(min(dates_scan_10d), max(dates_scan_10d), by = "5 years"), format = "%Y")
abline(v = stress_10d$stress_start, col = "firebrick",  lwd = 2,   lty = 2)
abline(h = stress_10d$stress_ES,    col = "firebrick",  lwd = 1.5, lty = 3)
abline(h = recent_ES_10d,           col = "darkorchid", lwd = 1.5, lty = 3)
legend("topright", legend = c("Rolling ES", "Stress start", "Stress ES", "Recent ES"),
       col = c("darkorange","firebrick","firebrick","darkorchid"),
       lwd = c(1,2,1.5,1.5), lty = c(1,2,3,3), bty = "n", cex = 0.75)

plot(dates_1d, returns_1d, type = "l", col = "steelblue", lwd = 0.6,
     main = "1-Day Returns - Stress Period Highlighted", xlab = "", ylab = "Return", xaxt = "n")
axis.Date(1, at = seq(min(dates_1d), max(dates_1d), by = "5 years"), format = "%Y")
rect(stress_1d$stress_start, par("usr")[3], stress_1d$stress_end, par("usr")[4],
     col = adjustcolor("firebrick", alpha.f = 0.15), border = NA)
abline(h = 0, lwd = 0.5)

plot(dates_10d, returns_10d, type = "l", col = "darkorange", lwd = 0.6,
     main = "10-Day Returns - Stress Period Highlighted", xlab = "", ylab = "Return", xaxt = "n")
axis.Date(1, at = seq(min(dates_10d), max(dates_10d), by = "5 years"), format = "%Y")
rect(stress_10d$stress_start, par("usr")[3], stress_10d$stress_end, par("usr")[4],
     col = adjustcolor("firebrick", alpha.f = 0.15), border = NA)
abline(h = 0, lwd = 0.5)

par(mfrow = c(1, 1))
```

## 7 Basel backtesting (real S&P 500)

The **primary** backtest: realised 1-day returns vs. the out-of-sample 99% VaR
computed above. Exceptions are counted, then the Basel **traffic light** is
evaluated over the most recent 250 trading days (the regulatory one-year window).

```{r backtest-real}
# Exception indicator on the REAL series (NA where VaR is absent -> FALSE)
breach_1d_vec <- (-returns_1d > VaR_1d) & !is.na(VaR_1d)
valid_idx_1d  <- which(!is.na(VaR_1d))

n_valid_1d  <- length(valid_idx_1d)
n_breach_1d <- sum(breach_1d_vec[valid_idx_1d])

# Basel traffic light over the most recent 250 trading days
tl_window       <- 250
recent_idx      <- tail(valid_idx_1d, tl_window)
recent_dates    <- dates_1d[recent_idx]
recent_breaches <- sum(breach_1d_vec[recent_idx])
tl_status <- if (recent_breaches <= 4) "GREEN" else if (recent_breaches <= 9) "AMBER" else "RED"

# Basel (1996) plus-factor: addition to the 'k' multiplier of 3
multiplier_table <- data.frame(
  breaches = 0:10,
  zone     = c(rep("GREEN", 5), rep("AMBER", 5), "RED"),
  addition = c(0.00, 0.00, 0.00, 0.00, 0.00, 0.40, 0.50, 0.65, 0.75, 0.85, 1.00)
)
k_addition <- multiplier_table$addition[multiplier_table$breaches == min(recent_breaches, 10)]

cat(sprintf("Full sample : %d / %d exceptions (%.2f%%, expected 1.00%%)\n",
            n_breach_1d, n_valid_1d, 100 * n_breach_1d / n_valid_1d))
cat(sprintf("Last 250 days: %s to %s\n", format(min(recent_dates)), format(max(recent_dates))))
cat(sprintf("              %d exceptions -> [ %s ]  (k add-on +%.2f)\n",
            recent_breaches, tl_status, k_addition))
```

```{r backtest-plot, fig.height=5}
valid_bt      <- !is.na(VaR_1d)
plot_dates    <- dates_1d[valid_bt]
plot_returns  <- returns_1d[valid_bt]
plot_VaR      <- VaR_1d[valid_bt]
plot_breaches <- breach_1d_vec[valid_bt]

plot(plot_dates, plot_returns, type = "l", col = "steelblue", lwd = 0.6,
     main = "S&P 500 - Realised 1-Day Returns vs. Rolling 99% VaR (out-of-sample)",
     xlab = "Date", ylab = "Daily return", xaxt = "n",
     ylim = range(c(plot_returns, -plot_VaR), na.rm = TRUE))
axis.Date(1, at = seq(min(plot_dates), max(plot_dates), by = "5 years"), format = "%Y")
lines(plot_dates, -plot_VaR, col = "firebrick", lwd = 1.4)
points(plot_dates[plot_breaches], plot_returns[plot_breaches], col = "red", pch = 19, cex = 0.5)
rect(min(recent_dates), par("usr")[3], max(recent_dates), par("usr")[4],
     col = adjustcolor("gold", alpha.f = 0.15), border = NA)
tl_col <- switch(tl_status, GREEN = "#2ecc71", AMBER = "#f39c12", RED = "#e74c3c")
legend("bottomleft",
       legend = c("1-day return", "-VaR 99%", "Exception",
                  sprintf("Last 250d (%s: %d exc.)", tl_status, recent_breaches)),
       col = c("steelblue", "firebrick", "red", adjustcolor("gold", 0.6)),
       pch = c(NA, NA, 19, 15), lwd = c(0.6, 1.4, NA, NA), pt.cex = c(NA, NA, 0.5, 1.5),
       bty = "n", cex = 0.75, text.col = c("black", "black", "black", tl_col))
abline(h = 0, col = "black", lwd = 0.4)
```

### Formal coverage tests — Kupiec & Christoffersen

Run on the **full-sample** 1-day exceptions ($N \approx 7{,}600$), where the
chi-square asymptotics are reliable. They are **not** run on the 10-day series:
overlapping returns make exceptions serially dependent by construction, which
violates the independence assumption. (On the short 250-day window the LR
asymptotics are also unreliable, so the zone above uses the Basel exact-binomial
cut-offs instead.)

```{r coverage-tests}
# Safe log-likelihood term: count*log(prob) with the convention 0*log(0) = 0
g <- function(count, prob) if (count == 0) 0 else count * log(prob)

kupiec_pof <- function(exceptions, p = 0.01) {
  exceptions <- exceptions[!is.na(exceptions)]
  N <- length(exceptions); x <- sum(exceptions); pi_hat <- x / N
  ll_null <- g(N - x, 1 - p)      + g(x, p)
  ll_alt  <- g(N - x, 1 - pi_hat) + g(x, pi_hat)
  LR <- -2 * (ll_null - ll_alt)
  list(N = N, x = x, rate = pi_hat, LR = LR, p_value = pchisq(LR, 1, lower.tail = FALSE))
}

christoffersen_ind <- function(exceptions) {
  ex <- as.integer(exceptions[!is.na(exceptions)]); n <- length(ex)
  prev <- ex[-n]; curr <- ex[-1]
  n00 <- sum(prev == 0 & curr == 0); n01 <- sum(prev == 0 & curr == 1)
  n10 <- sum(prev == 1 & curr == 0); n11 <- sum(prev == 1 & curr == 1)
  pi01 <- if ((n00 + n01) > 0) n01 / (n00 + n01) else 0
  pi11 <- if ((n10 + n11) > 0) n11 / (n10 + n11) else 0
  pi   <- (n01 + n11) / (n00 + n01 + n10 + n11)
  ll_null <- g(n00 + n10, 1 - pi)   + g(n01 + n11, pi)
  ll_alt  <- g(n00, 1 - pi01) + g(n01, pi01) + g(n10, 1 - pi11) + g(n11, pi11)
  LR <- -2 * (ll_null - ll_alt)
  list(LR = LR, p_value = pchisq(LR, 1, lower.tail = FALSE))
}

exc_1d <- as.integer(breach_1d_vec[valid_idx_1d])   # 0/1, contiguous in time
uc     <- kupiec_pof(exc_1d, p = 0.01)
ind    <- christoffersen_ind(exc_1d)
LR_cc  <- uc$LR + ind$LR
p_cc   <- pchisq(LR_cc, df = 2, lower.tail = FALSE)

cov_tbl <- data.frame(
  Test      = c("Kupiec POF (unconditional)", "Christoffersen (independence)", "Conditional coverage"),
  LR        = c(uc$LR, ind$LR, LR_cc),
  df        = c(1, 1, 2),
  `p-value` = c(uc$p_value, ind$p_value, p_cc),
  Decision  = c(ifelse(uc$p_value < 0.05, "reject", "do not reject"),
                ifelse(ind$p_value < 0.05, "reject", "do not reject"),
                ifelse(p_cc < 0.05, "reject", "do not reject")),
  check.names = FALSE
)
knitr::kable(cov_tbl, digits = 3,
             caption = sprintf("Backtest of 99%% VaR on %d observations, %d exceptions (%.2f%%).",
                               uc$N, uc$x, 100 * uc$rate))
```

On this sample the model is rejected on all three tests: the exception rate
exceeds 1% (Kupiec) **and** the exceptions cluster in time (Christoffersen
independence strongly rejected). Both are the textbook failure mode of
equal-weighted historical simulation — it reacts slowly to volatility changes
and does not capture volatility clustering, so breaches bunch together in crisis
periods. This is a known limitation of the method, not a coding artefact; a
conditional model (EWMA- or GARCH-filtered historical simulation) would target
both effects.

## 8 Appendix A — classifier sanity-check (synthetic, **not** an S&P result)

> This is **not** a backtest of the S&P 500 and **not** a result about the index.
> Its only purpose is to confirm that the breach-counting and zone logic behave
> correctly when fed a series with a known, deliberately stressed regime. The
> series is simulated from a regime-switching Normal model (no fat tails, no
> realistic clustering); the resulting zone depends on the RNG seed and is
> meaningless as a risk statement.

```{r appendix-synthetic, fig.height=5}
set.seed(42)

dates_syn <- seq(as.Date("2007-01-02"), Sys.Date(), by = "day")
dates_syn <- dates_syn[!weekdays(dates_syn) %in% c("Saturday", "Sunday")]
n_syn     <- length(dates_syn)

returns_syn <- numeric(n_syn)
for (i in seq_len(n_syn)) {
  d <- dates_syn[i]
  if (d >= as.Date("2008-09-01") & d <= as.Date("2009-06-30")) {
    returns_syn[i] <- rnorm(1, mean = -0.0015, sd = 0.028)   # GFC
  } else if (d >= as.Date("2020-02-20") & d <= as.Date("2020-05-31")) {
    returns_syn[i] <- rnorm(1, mean = -0.0008, sd = 0.025)   # COVID crash
  } else if ((d >= as.Date("2011-07-01") & d <= as.Date("2011-12-31")) |
             (d >= as.Date("2018-10-01") & d <= as.Date("2018-12-31")) |
             (d >= as.Date("2022-01-01") & d <= as.Date("2022-12-31"))) {
    returns_syn[i] <- rnorm(1, mean = 0.0000, sd = 0.016)    # elevated vol
  } else {
    returns_syn[i] <- rnorm(1, mean = 0.0003, sd = 0.010)    # normal
  }
}

VaR_syn        <- hist_VaR(returns_syn, window = 252, conf = 0.99)   # same OOS estimator
breach_syn_vec <- (-returns_syn > VaR_syn) & !is.na(VaR_syn)
valid_idx_syn  <- which(!is.na(VaR_syn))
recent_breaches_syn <- sum(breach_syn_vec[tail(valid_idx_syn, 250)])
tl_status_syn <- if (recent_breaches_syn <= 4) "GREEN" else if (recent_breaches_syn <= 9) "AMBER" else "RED"

cat(sprintf("Synthetic series: %d days | full-sample exceptions %.2f%% | last-250 zone: %s (%d)\n",
            n_syn, 100 * sum(breach_syn_vec[valid_idx_syn]) / length(valid_idx_syn),
            tl_status_syn, recent_breaches_syn))

plot_dates_syn <- dates_syn[valid_idx_syn]
plot(plot_dates_syn, returns_syn[valid_idx_syn], type = "l", col = "grey50", lwd = 0.5,
     main = "Appendix A - Synthetic series (classifier sanity-check, NOT S&P)",
     xlab = "Date", ylab = "Synthetic return", xaxt = "n",
     ylim = range(c(returns_syn[valid_idx_syn], -VaR_syn[valid_idx_syn]), na.rm = TRUE))
axis.Date(1, at = seq(min(plot_dates_syn), max(plot_dates_syn), by = "2 years"), format = "%Y")
lines(plot_dates_syn, -VaR_syn[valid_idx_syn], col = "firebrick", lwd = 1.2)
points(dates_syn[breach_syn_vec], returns_syn[breach_syn_vec], col = "red", pch = 19, cex = 0.4)
abline(h = 0, col = "black", lwd = 0.4)
legend("bottomleft", legend = c("synthetic return", "-VaR 99%", "exception"),
       col = c("grey50", "firebrick", "red"), pch = c(NA, NA, 19),
       lwd = c(0.5, 1.2, NA), pt.cex = c(NA, NA, 0.4), bty = "n", cex = 0.75)
```

## 9 Summary

On the realised index the 99% VaR averages
`r sprintf("%.2f", 100 * mean(VaR_1d, na.rm = TRUE))`% (1-day) and peaks at
`r sprintf("%.2f", 100 * max(VaR_1d, na.rm = TRUE))`% during the
`r format(stress_1d$stress_start)` – `r format(stress_1d$stress_end)` stress
window; the 97.5% ES averages
`r sprintf("%.2f", 100 * mean(ES_1d, na.rm = TRUE))`%. The simplified IMA-style
charge is bound by the stressed period
(`r sprintf("%.2f", 100 * IMA_1d)`% for 1-day,
`r sprintf("%.2f", 100 * IMA_10d)`% for 10-day). Backtesting the model on the
real returns gives `r uc$x`/`r uc$N` exceptions
(`r sprintf("%.2f", 100 * uc$rate)`%): unconditional coverage is rejected
(Kupiec *p* = `r sprintf("%.3f", uc$p_value)`) and so is independence
(Christoffersen *p* = `r sprintf("%.3f", ind$p_value)`) — i.e. breaches cluster,
the expected behaviour of equal-weighted historical simulation. Over the most
recent 250 days the model sits in the **`r tl_status`** zone
(`r recent_breaches` `r ifelse(recent_breaches == 1, "exception", "exceptions")`).

**Known limitations.**

- Equal-weighted historical-simulation VaR reacts slowly to volatility changes
  (ghosting; no volatility scaling) and does not model volatility clustering, so
  it under-reacts to regime shifts — visible here in both the above-1% exception
  rate and the rejected independence test (clustered breaches).
- The 10-day figures use **overlapping** returns and are therefore serially
  correlated; their exception counts are descriptive only, not comparable to 1%
  (so the formal coverage tests run on the 1-day series).
- The IMA-style charge is a **simplification** (no regulatory multiplier,
  liquidity horizons, NMRF/SES add-on, DRC, or cross-class aggregation).
- ES is reported via exceedance counts — a diagnostic, **not** a formal ES
  backtest (ES is not elicitable; see Acerbi–Székely 2014).

## Reproducibility

```{r sessioninfo}
sessionInfo()
```
